---
title: "statistical_learning"
otuput: github_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(glmnet)
set.seed(8105)
```


# LASSO


### Raw Data
```{r}
df <- read_csv("./data/birthweight.csv") 

head(df)
```



### Data Cleaning + Sampling

  * `as.factor()` : change dbl into fct, change variable type to use `fct_recode()`
  * `fct_recode(column, "new_value" = "old_value")` change factor level 
  * `as.logical()` change 0 or 1 into FALSE or TRUE
  * `sample_n(sample size)` sample from the cleaned data set

```{r,message=FALSE}
bwt_df = 
  read_csv("./data/birthweight.csv") %>% 
  janitor::clean_names() %>%
  mutate(
    babysex = as.factor(babysex),
    babysex = fct_recode(babysex, "male" = "1", "female" = "2"),
    frace = as.factor(frace),
    frace = fct_recode(frace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4", "other" = "8"),
    malform = as.logical(malform),
    mrace = as.factor(mrace),
    mrace = fct_recode(mrace, "white" = "1", "black" = "2", "asian" = "3", 
                       "puerto rican" = "4")) %>% 
  sample_n(200)  # get a sample of size 200

```



### Create Design Matrix

  * **glmnet** package predates tidyverse. It needs a design matrix (only contains X)


  * `model.matrix(formula = y ~ x, data = df)` can generate an expanded design matrix
by expanding factors to a set of dummy variables (depending on the contrasts) and expanding interactions similarly.

```{r}
x = model.matrix(bwt ~ ., bwt_df)[,-1]
x

y = bwt_df$bwt

```


```{r}
lambda = 10^(seq(3, -2, -0.1))

lasso_fit =
  glmnet(x, y, lambda = lambda)

lasso_cv =
  cv.glmnet(x, y, lambda = lambda)

lambda_opt = lasso_cv$lambda.min
```


```{r}
broom::tidy(lasso_fit) %>% 
  select(term, lambda, estimate) %>% 
  complete(term, lambda, fill = list(estimate = 0) ) %>% 
  filter(term != "(Intercept)") %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate, group = term, color = term)) + 
  geom_path() + 
  geom_vline(xintercept = log(lambda_opt, 10), color = "blue", size = 1.2) +
  theme(legend.position = "none")

```


```{r}
broom::tidy(lasso_cv) %>% 
  ggplot(aes(x = log(lambda, 10), y = estimate)) + 
  geom_point()  

```


```{r}
lasso_fit = 
  glmnet(x, y, lambda = lambda_opt)

lasso_fit %>% broom::tidy()

```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


```{r}


```


